---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 4.8,
  fig.align = "center",
  dpi = 144,
  out.width = "100%"
)
```

# gghf

`gghf` provides a small ggplot2 theme and colour scales aligned with
*Nasjonalt profilprogram for helseforetakene*.

The package is intentionally limited to plot styling primitives:

- `theme_hf()` for a clean ggplot2 theme
- `scale_colour_hf()` / `scale_color_hf()` for discrete colour scales
- `scale_fill_hf()` for discrete fill scales
- `hf_palette()` for direct access to official HEX colours

It does not include or recreate the protected prikkekors symbol, health trust
logos, or other trademarked sender marks.

## Installation

```r
pak::pak("lobot-de-bespin/gghf")
```

## Presentation

A short Quarto presentation is published with GitHub Pages:
<https://lobot-de-bespin.github.io/gghf/>

## Examples

```{r}
library("ggplot2")
library("gghf")

theme_set(theme_hf(base_family = "sans"))

service_levels <- c(
  "Akutt", "Barn", "Bildediagnostikk", "Føde", "Kirurgi", "Lab",
  "Medisin", "Nevro", "Psykisk helse", "Rehab", "Screening", "Teknologi"
)

activity <- data.frame(
  service = factor(service_levels, levels = service_levels),
  consultations = c(1280, 960, 1410, 820, 1160, 1540, 1320, 740, 1040, 680, 890, 570),
  change = c(7, 4, 9, -2, 6, 11, 5, -4, 3, -1, 8, 2)
)
```

### Full palette

```{r palette}
palette_df <- data.frame(
  name = factor(names(hf_palette()), levels = names(hf_palette())),
  colour = unname(hf_palette()),
  x = seq_along(hf_palette())
)

ggplot(palette_df, aes(x, 1, fill = name)) +
  geom_tile(width = 0.96, height = 0.82) +
  geom_text(aes(label = colour), colour = "white", fontface = "bold", size = 3) +
  scale_fill_hf(guide = "none") +
  scale_x_continuous(NULL, breaks = palette_df$x, labels = palette_df$name) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Official health trust profile colours") +
  theme_hf(base_family = "sans") +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid = element_blank()
  )
```

### Discrete fill scale

```{r activity-bars}
ggplot(activity, aes(reorder(service, consultations), consultations, fill = service)) +
  geom_col(width = 0.72) +
  scale_fill_hf(guide = "none") +
  coord_flip() +
  labs(
    title = "Activity by clinical area",
    subtitle = "All twelve profile colours are used before the palette repeats",
    x = NULL,
    y = "Consultations"
  )
```

### Discrete colour scale

```{r activity-change}
ggplot(activity, aes(consultations, change, colour = service)) +
  geom_hline(yintercept = 0, linewidth = 0.3, linetype = "dashed") +
  geom_point(size = 4) +
  scale_colour_hf() +
  labs(
    title = "Change in activity",
    subtitle = "The same palette works for colour aesthetics",
    x = "Consultations",
    y = "Change from previous period (%)",
    colour = "Clinical area"
  ) +
  theme(legend.position = "right")
```

### Small multiples

```{r small-multiples}
trend <- expand.grid(
  service = factor(service_levels[1:8], levels = service_levels[1:8]),
  month = factor(month.abb[1:6], levels = month.abb[1:6])
)
trend$value <- c(
  88, 91, 92, 95, 98, 101,
  72, 74, 75, 79, 78, 82,
  104, 106, 108, 111, 113, 117,
  64, 67, 65, 68, 70, 71,
  93, 92, 96, 99, 101, 104,
  112, 115, 119, 121, 125, 130,
  84, 87, 90, 91, 93, 96,
  58, 59, 61, 63, 62, 66
)

ggplot(trend, aes(month, value, group = service, colour = service)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(vars(service), ncol = 4) +
  scale_colour_hf(guide = "none") +
  labs(
    title = "Monthly trend by area",
    subtitle = "Facet labels and plot text use the health trust theme",
    x = NULL,
    y = "Index"
  )
```

## Source

Colour values and design direction are based on *Nasjonalt profilprogram for
helseforetakene*, final 2022 version, PDF metadata last modified 2025-02-26.
